---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(repr)
library(broom)
library(leaps)
library(moderndive)
library(gridExtra)
library(grid)
library(reshape2)
library(ggplot2)
library(lubridate)
```

```{r}
url <- "https://raw.githubusercontent.com/Wendy1907/Sleep_Efficiency_Project/main/Sleep_Efficiency.csv"
df <- read_csv(url)
head(df)
```


```{r}
df <- df %>% 
        rename("Wakeup.time" = "Wakeup time",
               "Sleep.duration" = "Sleep duration",
               "Sleep.efficiency" = "Sleep efficiency",
               "REM.sleep.percentage" = "REM sleep percentage",
               "Deep.sleep.percentage" = "Deep sleep percentage",
               "Light.sleep.percentage" = "Light sleep percentage",
               "Caffeine.consumption" = "Caffeine consumption",
               "Alcohol.consumption" = "Alcohol consumption",
               "Smoking.status" = "Smoking status",
               "Exercise.frequency" = "Exercise frequency"
              )
head(df)
```

```{r}
sum(is.na(df))
```

```{r}
colSums(is.na(df))
```

```{r}
df_clean <- df %>% drop_na(Awakenings, Caffeine.consumption, Alcohol.consumption, Exercise.frequency)
```

```{r}
sum(duplicated(df_clean))
```

```{r}
head(df_clean)
```

```{r}
data <- df_clean %>%
select(Caffeine.consumption, Alcohol.consumption, Smoking.status, Exercise.frequency, 
       Sleep.efficiency)
head(data)
```

```{r}
df_clean$Gender <- as.factor(df_clean$Gender) 
df_clean$Smoking.status <- as.factor(df_clean$Smoking.status)

#delete ID column because it doesn't carry any information
transformed_df <- df_clean %>% select(-ID)
# discard Bedtime and wakeup.time
transformed_df <- transformed_df %>% select(-Bedtime, -Wakeup.time)

#extract numeric columns only
numeric_df = transformed_df[, sapply(transformed_df, is.numeric)]
# calculate their correlation
cor_matrix <- cor(numeric_df, use = "complete.obs")
#plot heatmap
melted_cor_matrix <- melt(cor_matrix)
ggplot(melted_cor_matrix, aes(Var1, Var2, fill = value)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                        midpoint = 0, limit = c(-1,1), space = "Lab", 
                        name="Pearson\nCorrelation") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          axis.title = element_blank()) +
    coord_fixed()

```
Deep.sleep.percentage is highly correlated with Light.sleep.percentage. 
```{r}
transformed_df <- select(transformed_df, -Light.sleep.percentage)
```

```{r}
features <- c('Sleep.efficiency', 'Age', 'REM.sleep.percentage',
              'Deep.sleep.percentage', 
              'Awakenings', 'Caffeine.consumption', 'Alcohol.consumption',
              'Exercise.frequency')

plots = list()

for (feature in features) {
  p1 <- ggplot(transformed_df, aes_string(x = feature)) +
        geom_density(fill = "blue", alpha = 0.5) +
        labs(title = paste('Distribution of', feature))

  p2 <- ggplot(transformed_df, aes_string(y = feature)) +
        geom_boxplot() +
        labs(title = paste('Box Plot of', feature))
        coord_flip()

  plots[[length(plots) + 1]] <- p1
  plots[[length(plots) + 1]] <- p2
}
```

```{r}
#index into the plots list to check distribution of variables
do.call(grid.arrange, c(plots[1:2]))
do.call(grid.arrange, c(plots[11:12]))
do.call(grid.arrange, c(plots[13:14]))
do.call(grid.arrange, c(plots[15:16]))
```
Sleep.efficiency is left-skewed.
Caffeine consumption is right-skewed
Alcohol consumption is right-skewed
exercise frequency is right-skewd


```{r}
table(transformed_df$Smoking.status)
```


Because we want to remove the effects of confouding variables, we fit a multivariate
linear regression using the full transformed dataset. Noted that sleep efficiency
is a percentage between 0 and 1. We want to make sure the fitted values are within 
the range of 0 and 1.
```{r}
FullModel = lm(Sleep.efficiency~.,data = transformed_df)
summary(FullModel)
```

do variable selection
```{r}
reg = regsubsets(Sleep.efficiency~., data = transformed_df, nvmax = 10)
s_reg = summary(reg)
cp_values = s_reg$cp
num_predictors = seq(along = cp_values)
plot(num_predictors, cp_values, type = "b", xlab = "Number of Predictors", 
     ylab = "Cp", main = "Cp vs p")
abline(a=0,b=1)
```
choose p = 8.
```{r}
print(s_reg$which)
```

```{r}
transformed_df = transformed_df %>% select(-Gender, -Sleep.duration)
bestModel = lm(Sleep.efficiency~., data = transformed_df)
summary(bestModel)
kappa(bestModel)
plot(bestModel)
```

```{r}
which(abs(rstudent(bestModel))>3) ## No outliers
which(cooks.distance(bestModel)>0.5) ## No influential points
```

```{r}

boxplot(Sleep.efficiency~Smoking.status, data = transformed_df, 
        main = "Boxplot By smoking status")
```

```{r}
pca_raw = transformed_df %>% select(Caffeine.consumption, Alcohol.consumption, Exercise.frequency)
pca_result = prcomp(pca_raw, scale= TRUE, center = TRUE)
# use the first two principal components to visual sleep efficiency
proj_pca <- predict(pca_result)[,1:2] |>
  as_tibble() |>
  mutate(Sleep.efficiency = transformed_df$Sleep.efficiency)

ggplot(proj_pca, aes(x = PC1, y = PC2, color = Sleep.efficiency)) + 
  geom_point() +
  scale_color_gradient(low = "blue", high = "red")


scaled_raw = scale(pca_raw,center = TRUE, scale=TRUE)
proj_pca = data.frame(proj_pca, scaled_raw)
proj_pca = pivot_longer(proj_pca, cols = names(pca_raw), names_to = "features", 
                      values_to = "score")

ggplot(proj_pca, aes(x= PC1, y=PC2, color = score)) +
  geom_point() +
  scale_color_gradient(low = "blue", high="red") +
  facet_wrap(~features, ncol = 3)


pca_result$rotation[, 1:2] |>
  as_tibble() |>
  mutate(feature = names(pca_raw)) |>
  pivot_longer(-feature) |>
  ggplot(aes(value, feature, fill = feature)) +
  facet_wrap(~name) +
  geom_col() +
  theme(legend.position = "none", axis.title = element_blank()) +
  geom_vline(xintercept = 0)

```